### 2021-07-29 doc/thesis/chapters/deployment-on-hardware.tex
Fig. -> Figure
incorporate some of the feedback (wip)

The USRP b200mini is a device that can be used to create a software-defined radio.
It resembles the size of a business card and is developed by Ettus Research.
A signal range from 70 MHz to 6 GHz can be covered by the front end.
The board further features a Xilinx Spartan-6 FPGA that is user-programmable, which makes the device very flexible for various applications with the received signal \cite{USRPb200Mini}.\\

@@

The block diagram in Fig.\ref{fig:bd_impl_vivado} shows the implemented architecture that is used in the SoC.
The various design domains in and around the SoC are depicted in different visual representations.
The FPGA domain blocks are highlighted in gray, with the main IPs being surrounded by emphasized lines.
The processor system domain is shown as white blocks.
Blocks that are external to the SoC are drawn with dashed lines.

>>>>

The USRP b200mini is a device that can be used to create a software-defined radio.
It resembles the size of a business card and is developed by Ettus Research.
A signal range from 70~MHz to 6~GHz can be covered by the front end.
The board further features a Xilinx Spartan-6 FPGA that is user-programmable, which makes the device very flexible for various applications with the received signal \cite{USRPb200Mini}.\\

@@

The block diagram in Figure \ref{fig:bd_impl_vivado} shows the implemented architecture that is used in the SoC.
The various design domains in and around the SoC are depicted in different visual representations.
The FPGA domain blocks are highlighted in gray, with the main IPs being surrounded by emphasized lines.
The processor system domain is shown as white blocks.
Blocks that are external to the SoC are drawn with dashed lines.

### 2021-07-29 doc/thesis/chapters/dsp.tex
incorporate some of the feedback (wip)

Wireless transmission systems usually transmit their signals over the air.
Various frequency bands and different types of modulation are being used for this purpose.
The choice of frequency, as well as the choice of the modulation type, have impacts on the transmission characteristics.
Depending on these choices, positive effects can be achieved in the amount of data that can be sent over the transmission channel, for example.
Another advantage is the ability to adapt the ongoing transmission to the changing characteristics of the channel.
This is especially useful for mobile receivers, since the signals' path is inherently changing over time there, which leads to challenging conditions for the receiver.
Additionally, varying signal strength, or multipath propagation may occur.
Multipath propagation is caused by reflections of the signal, which - for example - can occur on a house facade, or a large vehicle that is passing by.
The receiver then gets a signal, that is a summation of the direct path and the reflected path - a signal from multiple paths.
Besides that, environmental impacts, such as rain or fog have an effect on the transmission channel and thus the received signal.\\

Many more factors could be listed, and elaborations over the various modulation types fill entire books.
However, with all these different facts, there are things that they all have in common.
On the transmitter side, all these signals are transmitted in a band-limited, high frequency signal.
The band limitation needs to be given at the transmitting antenna, to minimize interference to other bands in the frequency spectrum.
The frequency needs to be high, to radiate the signal by an antenna.
On the receiver side, once the signal reaches the antenna of a receiver, there are further similarities.
The antenna signal needs to be amplified and filtered at first.
Both steps depend on the antenna characteristics, but usually the received signal is of relatively low power, and the antenna receives a wider range of frequencies than required.
The filter therefor selects the frequency area that is of interest.
Another reason for the filter is to avoid spectral replicas in the subsequent modulation, or down-conversion.
Here, the front-end often down-converts to a low intermediate frequency (IF), instead of directly going to baseband.
The intermediate frequency is then shifted down to baseband by a modulator later in the digital domain, which brings advantages in the downstream processing chain. %TODO check this or explain more which advantages (HF to BB directly --> i/q complex signal? ; HF to IF to BB --> real signal ) ??
These three steps - amplification, filtering and down-conversion to an IF - are often done in the analog domain, since digital circuits cannot work with such high frequencies.
After the analog signal is down-converted to baseband, it needs to be digitized.
This is done by an analog-digital-converter (ADC).
After the signal is processed to digital baseband, the received signal can finally be decoded to get the actual data content.
This decoding process depends on the respective modulation type that is used.\\

@@

Multiple different receiver architectures exist and are implemented in receivers.
Two common examples are explained in the following sections.

@@

Figure \ref{fig:bd_superhet} shows a block diagram of a superhet receiver.
The antenna signal first runs through a band selection filter.
It selects the desired frequency band and filters out any unwanted signals outside of this band.
A low-noise amplifier (LNA) may be used to boost the signal strenght.
The next step is the mixer, which is fed by a local oscillator (LO).
This shifts, or down-converts, the antenna signal down to the IF.
Afterwards, the IF signal is amplified and filtered again to remove the side-products of the mixer, namely the image replicas.
Thus, it is also called an image rejection filter.
The demodulator stage can now perform any further signal processing on the signal, until the required signal is retrieved \cite{PassosFábio2020AHSo}.\\

The depicted block diagram may represent an FM broadcast receiver, where an audio output is generated from the antenna input.

@@

A direct-conversion receiver, as its name suggests, converts an RF signal to baseband directly.
It can also be called homodyne, synchrodyne, or zero-IF receiver.\\

In order to achieve direct down-conversion, the local oscillator of the mixer needs to be set exactly to the center frequency of the desired signal.
The main advantage of direct-conversion receivers is, that their output is an image-free signal.
This is due to the properties of the mixer.
The strongest disadvantages are the LO leakage, which leads to a DC offset at the mixer output, and the flicker noise, also called \textit{1/f}-noise, which induces strong noise at low frequencies \cite{PassosFábio2020AHSo}.
These effects are not explained into more detail here, since this is not the focus of this thesis.

@@

The procedure that is used for down-conversion is an amplitude modulation (AM).
The equation for a generic AM signal is $ S_{AM}(t) = A_m m(t) \cdot A_c c(t) $, where $m(t)$ represents the message signal, and $c(t)$ is the high-frequency carrier.
Their respective amplitudes are $A_m$ and $A_c$.
This equation is depicted as a block diagram and signal spectrum in Figure \ref{fig:bd_down_conversion_spectrum}, where $S_{AM}$ is denoted as the baseband signal $m_{BB}$ of the original signal $m$.
The signal $m(t)$ is a baseband signal.
It may be antenna signal that was filtered by a bandpass filter.
The received signal is a real signal, and thus its negative frequency spectrum is an exact mirror of the positive frequencies.
This fact can be explained in two simple equations.\\

It is well-known that the real signal $cos(wt) = \frac{1}{2} \Big( e^{jw} + e^{-jw} \Big)$, which follows from $e^{jw} = cos(wt)+jsin(wt)$.
Looking at the frequencies of the first equation, the negative and positive parts can be seen.\\

@@

Assuming that $m(t) = cos(\omega_mt)$, $c(t) = cos(\omega_ct)$ and $A_m = A_c = 1$,

@@

follows.
By applying the mathematical theorem

@@

the final formula of the conversion can be represented as

@@

In words: The sampling frequency must be higher than twice the frequency of the signals' highest frequency content.
Any digital signal processing technique must hold the specification of the sampling theorem \cite[chpt. 4.2]{AlessioSilviaMaria2016DSPa}.
In case the theorem is violated, the effect of aliasing occurs on the generated signal.
The sampling process introduces spectral replicas at integer multiples of the sampling frequency $f_s$.
These replicas overlap, if $f_s$ does not meet the theorem, which results in signal distortion.
Figure \ref{fig:aliasing_freq_spectrum} illustrates the effect of aliasing in the frequency spectrum \cite{ThyagarajanK.S2019ItDS}.

\includepicture [0.95] {
  Frequency spectrum of the sampled baseband signal with its replicas at integer multiples of $f_s$ \cite{ThyagarajanK.S2019ItDS}.\\
  A) $f_s > 2f_{signal}$: $f_s$ is high enough and thus meets the theorem.\\
  B) $f_s = 2f_{signal}$: $f_s$ exactly meets the theorem.\\
  C) $f_s < 2f_{signal}$: $f_s$ is too low and thus creates overlaps of the replicas - aliasing.
} {aliasing_freq_spectrum} {img/draw.io/sampling-aliasing}

@@

As an example, assuming the signal of interest has a bandwidth of 20 kHz.
The Nyquist theorem thus requires a remaining sampling rate of at least twice the signals' frequency, so 40 kHz.
The signal is sampled with an oversampling factor of 10 - meaning a sample rate of 200 kHz.
A downsampling factor of 4 is now applied, which results in a sample rate of 50 kHz.
The resulting signal is now downsampled and can be processed at this lower data rate, while none of the original data content is lost.\\

@@

More detailed information can be found in \cite[chpt. 6.9]{AlessioSilviaMaria2016DSPa}, \cite[chpt. 4.1]{GopiE.S2018MDSP} and \cite[chpt. 10.2.2]{ThyagarajanK.S2019ItDS}.

@@

The instantanious frequency $f_i$ of an FM signal can be described as
\begin{equation}
  f_i = f_c + \Delta f \cdot m(t)
\end{equation}
where $f_c$ is the carrier- or center frequency, $\Delta f$ is the maximum frequency deviation and $m(t)$ is the information or message signal that is to be transmitted.
Simply looking at this equation, the frequency deviation, which is sometimes also called the swing, varies in the range between $f_c \pm \Delta f$.
Considering the relationship between frequency and angle, as described above, and after some simple substitutions which will not be described into detail here, the equation for a generic frequency modulated wave is the following
\begin{equation}
  s(t) = A_c\ cos \Big( 2 \pi f_c t + 2 \pi k_f \int m(t) dt \Big)
  \label{equ:fm_func}
\end{equation}
where $A_c$ is the amplitude of the resulting FM signal and $k_f$ is the frequency sensitivity factor.
This sensitivity factor has a direct relationship with the modulation index $\beta$.
\begin{equation}
  \beta = \frac{\Delta f}{f_m} = \frac{k_f A_m}{f_m}
\end{equation}
where $f_m$ is the highest existing frequency, or the bandwidth of the information signal.
% More info on modulation index: https://fas.org/man/dod-101/navy/docs/es310/FM.htm
Formula \eqref{equ:fm_func} describes a frequency modulated signal with a generic message signal $m(t)$.
A widely used example application of FM is broadcast radio, where audio streams are transmitted.
An audio stream can be described as a cosine wave.
Therefore, the information signal that is to be transmitted in broadcast radio can be represented as a cosine wave in the form of
\begin{equation}
  m(t) = A_m\ cos(2 \pi f_m t).
\end{equation}
By inserting this message signal into the generic FM formula \eqref{equ:fm_func}, the final equation for an FM signal transforms into the following form.
\begin{equation}
  y(t) = A_c\ cos \Big(2 \pi f_c t + \beta\ sin(2 \pi f_m t)\Big )
\end{equation}
Several formulas, as well as their derivations are described according to \cite[pg.54-55]{MatinMohammadA2018CSfE}.

@@

The frequency band that is used for FM broadcasting is defined worldwide and spans from 87.5 to 108 MHz.
However, some countries only partially use the band \cite[RR5-39]{ItuRadioRegulations}.
The range is located within the so-called Very-High-Frequency (VHF) band, which is open to the public in terms of usage.
This means, that any transmitter may use the specified frequency range freely.
Austria allowed the legal usage in 2006 \cite{AustriaRundfunkgesetz2014}.
However, the transmission power needs to be limited, so that neighboring receivers are not disturbed in receiving any existing channels.
The European Commision specifies the power limit as 50 nW of effective radiated power (ERP) \cite{EuCommissionRadioSpectrumHarmonization}.
Another limitation is, that a transmitter must be capable to transmit on multiple center frequencies within the broadcasting band.
It must not be fixed to a single center frequency.
Through this specification, a transmitter is capable of switching its transmission center frequency, in order to avoid interference with another transmitters signal.\\

@@

The entire FM broadcasting band from 87.5 to 108 MHz divides into multiple channels which can be used.
In Europe, the European Telecommunications Standards Institute (ETSI) sets the respective standards for the usage of these frequencies.
The main specifications for a single FM broadcasting channel are explained in this section.\\

Each channel may allocate a bandwidth of 200 kHz.
Within one of these channels, the maximum deviation from the center frequency shall not exceed 75 kHz.
This leaves a guard band of 25 kHz on either side, to minimize interference with adjacent channels.
Center frequencies may be allocated on multiples of 100 kHz.
However, neighboring channels need to be considered therefore, since this may result in an overlap in the spectrum.\\

Figure \ref{fig:fm_channel_baseband_freqs} shows the frequency spectrum of a single channel, located at its carrier- or center frequency.
For a better overview, only the positive spectum part is shown.
The negative part is an exact mirror, because the transmitted and received signals from the antenna are real signals - they do not have an imaginary part.
The spectrum consists of multiple parts. Therefore, it is also referred to as `multiplex signal' in the literature. Several parts are described in the following paragraphs \cite{EbuFmRegulators}\cite{EtsiFmSoundBroadcastingServiceHarmonizedStandard}\cite{ElectronicsNotesFmTutorial}.

@@

The mono audio stream is located between 30 Hz and 15 kHz.
This signal is built by the sum of the left and right audio channels.
%The lower limit at 30 Hz is to prevent the transmission of a direct current (DC) part, which would require a large amount of power in transmission.
%This effect is one of the downsides of amplitude modulation (AM), where the carrier is located at the channels' center frequeny and requires a high transmission power. % TODO: check this fact
The upper limit at 15 kHz is chosen to maintain a sufficient spacing to the first subcarrier.
For audio streaming, as in FM broadcasting, this is not a limitation, since it is unlikely to have frequencies higher than 15 kHz in an audio stream.
Also, this already reaches the upper limit of the human ears' bandwidth, so higher spectral parts will not be heard anyway.

@@

The first subcarrier is allocated at an offset of 19 kHz from the center frequency.
This subcarrier is also called the 'pilot tone', since it is a continuous signal.
It is independent of any varying message signal content.
The pilot tone is used for stereo audio demodulation.
To regenerate a stereo audio signal, the left and right audio channels need to be recovered correctly.

@@

A signal that is constructed by the difference between the left and right audio channel is modulated on a 38 kHz subcarrier.
This subcarrier is an integer multiple of the pilot tone at 19 kHz, for practical reasons.
However, the 38 kHz carrier is suppressed and thus not visible in the received spectrum.
The modulation technique that is used for this spectral part is called dual-sideband suppressed-carrier (DSB-SC).
Even though the carrier is suppressed, it can still be recovered, because it is phase coherent with the 19 kHz pilot tone per definition.\\

The bandwidth for this difference-signal spans 15 kHz on either side of the subcarrier.
It is used to generate a stereo audio signal, in combination with the mono signal.
This process is explained into more detail in chapter \ref{sec:demodulation-of-broadcast-fm}.

@@

Considering all these parts in the spectrum, there is still free bandwidth available to use up to the maximum channel bandwitdh of 100 kHz in one sideband.
Because of that, additional services were added to the pure audio transmission, to provide additional data services and information.\\

Services that were implemented are the Data Radio Channel (DARC), which is mostly used in Japan and the USA, the Subsidary Communication Authorization (SCA) and the Radio Data System (RDS) \cite{RohdeSchwarzFM}.
Out of these, RDS is the most significant service in Europe.
It is used to transmit additional information about the channel, such as the radio stations' name, the currently playing songs' title or traffic information.

@@

The differentiator already converts the FM signal into an AM signal, but still leaves an FM content.
The transformation that happens here can be expressed in a formula, by simply differentiating Eq.\eqref{equ:fm_func}.
This operation, with some additional simple algebraic transformations, results in
\begin{equation}
  \frac{d s(t)}{dt} = A_c 2 \pi \Big[f_c + k_f m(t) \Big] sin \Big(\omega_c t + 2 \pi k_f \int m(t) dt -pi \Big)
  \label{equ:fm_demod_discriminator}
\end{equation}

This formula shows the AM and an FM contents clearly.
The FM part is described by the sine function, with its argument.
The more important part is the AM part, which resembles the message signal.
It is described by the term within the square brackets.

@@

An important requirement for the differentiator is, that the input signal is of a constant amplitude \cite{SchnyderHaller2002}.
The transmitted signal is subject to random additive noise on the transmission channel.
The differentiator would deliver wrong results because of this noise, when substracting two consecutive samples, as described in Eq.\eqref{equ:differentiator_two_samples}.
To achieve a constant amplitude, the received complex baseband signal $s(t)$ can be normalized to a value of one.
\begin{equation}
  s_{norm} = \frac{s}{|s|} = \frac{A(n)\ e^{j\phi_{FM}(n)}}{|{A(n)\ e^{j\phi_{FM}(n)}|}} = e^{j\phi_{FM}(n)}
\end{equation}
\begin{equation}
  e^{j\phi_{FM}(n)} = |1|
\end{equation}

@@

The differentiated signal still consists of a FM part, as explained above, using Eq.\eqref{equ:fm_demod_discriminator}.
In order to remove this high-frequeny part and extract the low-frequency envelope, the signal needs to be fed into an envelope detector.
A block diagram is shown in Figure \ref{fig:bd_freq_discriminator}.\\

@@

The first part of the channel frequency spectrum is the mono audio signal.
It is the summation signal of the left and right audio channel.
Thus, it contains enough information to replay an audio stream in mono quality.
A mono receiver can thus simply replay the so-called multiplex signal that is produced by the FM demodulator, without any further filtering or processing.
The 19 kHz pilot tone, and any higher frequency components, will not be audible by most people, because it is outside of the range of a human ear.
Besides that, any used speaker may also be unable to produce a frequency in this range.\\

Consequently, in order to demodulate this spectral part, a lowpass filter is required.
The filter needs to have a cutoff frequency of 15 kHz, to allow the entire mono audio spectrum to pass.
Sufficient attenutation must be available before 19 kHz, in order to suppress the pilot tone.

@@

The block diagram in Figure \ref{fig_bd_stereo_demod} illustrates the signal processing chain for stereo audio.
In the block diagram, $x(t)$ represents the multiplex signal from the FM demodulator.
The signals $x_l(t)$ and $x_r(t)$ describe the left and right audio signal, respectively.\\

@@

The central branch selects the pilot tone at 19 kHz with a bandpass filter.
This bandpass needs to have a frequency response that is sharp enough to have a sufficient attenuation at $\pm$4 kHz, since this is where the mono and difference signals' frequency spectra are allocated.
Afterwards, the pilot tone frequency is doubled to achieve a 38 kHz subcarrier.
This guarantees a phase-coherency between pilot and subcarrier, which is required to correctly shift the difference signal down to baseband.\\

The frequency doubling can be implemented with multiple methods.
The pilot tone can be modulated by simply multiplying it with itself.
Another option is to use a PLL and configure it to generate the double frequency at its output.
The PLL lock indicator can then be re-used as an indicator, whether a pilot tone is existent.\\

In the upper branch, a bandpass filter selects the difference signal with ranges from 23 to 53 kHz.
It is then modulated to baseband by the previously generated 38 kHz subcarrier.
Subsequently, a lowpass limits signal bandwidth to 15 kHz, which removes the generated modulation artifacts.\\

@@

The rightmost part in the diagram performs the combination of the mono and the difference signal, according to Eq.\eqref{equ_stereo_from_sum_diff}.

>>>>

Wireless transmission systems usually transmit their signals over the air.
Various frequency bands and different types of modulation are being used for this purpose.
The choice of frequency, as well as the choice of the modulation type, have impacts on the transmission characteristics.
Depending on these choices, positive effects can be achieved in the amount of data that can be sent over the transmission channel, for example.
Another advantage is the ability to adapt the ongoing transmission to the changing characteristics of the channel.
This is especially useful for mobile receivers, since the signals' path is inherently changing over time, which leads to challenging conditions for the receiver.
Additionally, varying signal strength, or multipath propagation may occur.
Multipath propagation is caused by reflections of the signal, which, for example, can occur on a house facade, or a large vehicle that is passing by.
The receiver then gets a signal that is a summation of the direct path and the reflected path -- a signal from multiple paths.
Besides that, environmental impacts, such as rain or fog have an effect on the transmission channel and thus the received signal.\\

Many more factors could be listed, and elaborations over the various modulation types fill entire books.
However, with all these different facts, there are things that they all have in common.
On the transmitter side, all these signals are transmitted in a band-limited, high frequency signal.
The band limitation needs to be given at the transmitting antenna, to minimize interference to other bands in the frequency spectrum.
The frequency needs to be high, to radiate the signal by an antenna.
On the receiver side, once the signal reaches the antenna of a receiver, there are further similarities.
The antenna signal needs to be amplified and filtered at first.
Both steps depend on the antenna characteristics, but usually the received signal is of relatively low power, and the antenna receives a wider range of frequencies than required.
The filter therefor selects the frequency area that is of interest.
Another reason for the filter is to avoid spectral replicas in the subsequent modulation, or down-conversion.
Here, the front-end often down-converts to a low intermediate frequency (IF), instead of directly going to baseband.
The intermediate frequency is then shifted down to baseband by a modulator later in the digital domain, which brings advantages in the downstream processing chain. %TODO check this or explain more which advantages (HF to BB directly --> i/q complex signal? ; HF to IF to BB --> real signal ) ??
These three steps -- amplification, filtering and down-conversion to an IF -- are often done in the analog domain, since digital circuits cannot work with such high frequencies.
After the analog signal is down-converted to baseband, it needs to be digitized.
This is done by an analog-digital-converter (ADC).
After the signal is processed to digital baseband, the received signal can finally be decoded to get the actual data content.
This decoding process depends on the respective modulation type that is used.\\

@@

Multiple different receiver architectures exist and are implemented in receivers.
Two common examples are explained in the following sections.
The main source of information for this section is \cite{PassosFábio2020AHSo}.

@@

Figure \ref{fig:bd_superhet} shows a block diagram of a superhet receiver.
The antenna signal first runs through a band selection filter.
It selects the desired frequency band and filters out any unwanted signals outside of this band.
A low-noise amplifier (LNA) may be used to boost the signal strenght.
The next step is the mixer, which is fed by a local oscillator (LO).
This shifts, or down-converts, the antenna signal down to the IF.
Afterwards, the IF signal is amplified and filtered again to remove the side-products of the mixer, namely the image replicas.
Thus, it is also called an image rejection filter.
The demodulator stage can now perform any further signal processing on the signal, until the required signal is retrieved.\\

The depicted block diagram represents a superhet receiver structure, for example an FM broadcast receiver, where an audio output is generated from the antenna input.

@@

A direct-conversion receiver, as its name suggests, converts an RF signal directly to baseband.
It can also be called homodyne, synchrodyne, or zero-IF receiver.\\

In order to achieve direct down-conversion, the local oscillator of the mixer needs to be set exactly to the center frequency of the desired signal.
The main advantage of direct-conversion receivers is, that their output is an image-free signal.
This is due to the properties of the mixer.
The strongest disadvantages are the LO leakage, which leads to a DC offset at the mixer's output, and the flicker noise, also called \textit{1/f}-noise, which induces strong noise at low frequencies.
These effects are not explained into more detail here, since this is not the focus of this thesis.

@@

The procedure that is used for down-conversion is an amplitude modulation (AM).
The equation for a generic AM signal is $ S_{AM}(t) = A_m m(t) \cdot A_c c(t) $, where $m(t)$ represents the message signal, and $c(t)$ is the high-frequency carrier.
Their respective amplitudes are $A_m$ and $A_c$.
This equation is depicted as a block diagram and signal spectrum in Figure \ref{fig:bd_down_conversion_spectrum}, where $S_{AM}$ is denoted as the baseband signal $m_{BB}$ of the original signal $m$.
The signal $m(t)$ is a baseband signal.
It may be antenna signal that was filtered by a bandpass filter.
The received signal is a real signal, and thus, its negative frequency spectrum is an exact mirror of the positive frequencies.
This fact can be explained in two simple equations.\\

It is well-known that the real signal $cos(wt) = \frac{1}{2} \Big( e^{jw} + e^{-jw} \Big)$, which follows from $e^{jw} = cos(wt)+jsin(wt)$.
Looking at the frequencies $\omega$ of the first equation, the negative and positive parts can be seen.\\

@@

\noindent
Assuming that $m(t) = cos(\omega_mt)$, $c(t) = cos(\omega_ct)$ and $A_m = A_c = 1$,

@@

\noindent
follows.
By applying the mathematical theorem

@@

\noindent
the final formula of the conversion can be represented as

@@

In words: The sampling frequency must be higher than twice the frequency of the signal's highest frequency content.
Any digital signal processing technique must hold the specification of the sampling theorem \cite[Ch. 4.2]{AlessioSilviaMaria2016DSPa}.
In case the theorem is violated, the effect of aliasing occurs on the generated signal.
The sampling process introduces spectral replicas at integer multiples of the sampling frequency $f_s$.
These replicas overlap, if $f_s$ does not meet the theorem, which results in signal distortion.
Figure \ref{fig:aliasing_freq_spectrum} illustrates the effect of aliasing in the frequency spectrum \cite{ThyagarajanK.S2019ItDS}.

\includepicture [0.95] {
  Frequency spectrum of the sampled baseband signal with its replicas at integer multiples of $f_s$ \cite{ThyagarajanK.S2019ItDS}.\\
  A) $f_s > 2f_{signal}$: $f_s$ is high enough, and thus, meets the theorem.\\
  B) $f_s = 2f_{signal}$: $f_s$ exactly meets the theorem.\\
  C) $f_s < 2f_{signal}$: $f_s$ is too low, and thus, creates overlaps of the replicas -- aliasing.
} {aliasing_freq_spectrum} {img/draw.io/sampling-aliasing}

@@

As an example, assuming the signal of interest has a bandwidth of 20~kHz.
The Nyquist theorem thus requires a remaining sampling rate of at least twice the signals' frequency, so 40~kHz.
The signal is sampled with an oversampling factor of 10 -- meaning a sample rate of 200~kHz.
A downsampling factor of 4 is now applied, which results in a sample rate of 50~kHz.
The resulting signal is now downsampled and can be processed at this lower data rate, while none of the original data content is lost.\\

@@

More detailed information can be found in \cite[Ch.~6.9]{AlessioSilviaMaria2016DSPa}, \cite[Ch.~4.1]{GopiE.S2018MDSP} and \cite[Ch.~10.2.2]{ThyagarajanK.S2019ItDS}.

@@

The instantanious frequency $f_i$ of an FM signal can be described as
\begin{equation}
  f_i = f_c + \Delta f \cdot m(t)
\end{equation}
where $f_c$ is the carrier- or center frequency, $\Delta f$ is the maximum frequency deviation and $m(t)$ is the information or message signal that is to be transmitted.
Simply looking at this equation, the frequency deviation, which is sometimes also called the swing, varies in the range between $f_c \pm \Delta f$.
Considering the relationship between frequency and angle, as described above, and after some simple substitutions which will not be described into detail here, the equation for a generic frequency modulated wave is the following
\begin{equation}
  s(t) = A_c\ cos \Big( 2 \pi f_c t + 2 \pi k_f \int m(t) dt \Big),
  \label{equ:fm_func}
\end{equation}
where $A_c$ is the amplitude of the resulting FM signal and $k_f$ is the frequency sensitivity factor.
This sensitivity factor has a direct relationship with the modulation index $\beta$.
\begin{equation}
  \beta = \frac{\Delta f}{f_m} = \frac{k_f A_m}{f_m},
\end{equation}
where $f_m$ is the highest existing frequency, or the bandwidth of the information signal.
% More info on modulation index: https://fas.org/man/dod-101/navy/docs/es310/FM.htm
Formula \eqref{equ:fm_func} describes a frequency modulated signal with a generic message signal $m(t)$.
A widely used example application of FM is broadcast radio, where audio streams are transmitted.
An audio stream can be described as a cosine wave.
Therefore, the information signal that is to be transmitted in broadcast radio can be represented as a cosine wave in the form of
\begin{equation}
  m(t) = A_m\ cos(2 \pi f_m t).
\end{equation}
By inserting this message signal into the generic FM Equation \eqref{equ:fm_func}, the final formula for an FM signal transforms into the following form.
\begin{equation}
  y(t) = A_c\ cos \Big(2 \pi f_c t + \beta\ sin(2 \pi f_m t)\Big )
\end{equation}
Several formulas, as well as their derivations are described according to \cite[pg.54-55]{MatinMohammadA2018CSfE}.

@@

The frequency band that is used for FM broadcasting is defined worldwide and spans from 87.5 to 108~MHz.
However, some countries only partially use the band \cite[RR5-39]{ItuRadioRegulations}.
The range is located within the so-called Very-High-Frequency (VHF) band, which is open to the public in terms of usage.
This means, that any transmitter may use the specified frequency range freely.
Austria allowed the legal usage in 2006 \cite{AustriaRundfunkgesetz2014}.
However, the transmission power needs to be limited, so that neighboring receivers are not disturbed in receiving any existing channels.
The European Commision specifies the power limit as 50 nW of effective radiated power (ERP) \cite{EuCommissionRadioSpectrumHarmonization}.
Another limitation is, that a transmitter must be capable to transmit on multiple center frequencies within the broadcasting band.
It must not be fixed to a single center frequency.
Through this specification, a transmitter is capable of switching its transmission center frequency, in order to avoid interference with another transmitters signal.\\

@@

The entire FM broadcasting band from 87.5 to 108~MHz divides into multiple channels which can be used.
In Europe, the European Telecommunications Standards Institute (ETSI) sets the respective standards for the usage of these frequencies.
The main specifications for a single FM broadcasting channel are explained in this section.\\

Each channel may allocate a bandwidth of 200~kHz.
Within one of these channels, the maximum deviation from the center frequency shall not exceed 75~kHz.
This leaves a guard band of 25~kHz on either side, to minimize interference with adjacent channels.
Center frequencies may be allocated on multiples of 100~kHz.
However, neighboring channels need to be considered therefore, since this may result in an overlap in the spectrum.\\

Figure \ref{fig:fm_channel_baseband_freqs} shows the frequency spectrum of a single channel, located at its carrier- or center frequency.
For a better overview, only the positive spectum part is shown.
The negative part is an exact mirror, because the transmitted and received signals from the antenna are real signals - they do not have an imaginary part.
The spectrum consists of multiple parts.
Therefore, it is also referred to as `multiplex signal' in the literature.
Several parts are described in the following paragraphs \cite{EbuFmRegulators},\cite{EtsiFmSoundBroadcastingServiceHarmonizedStandard},\cite{ElectronicsNotesFmTutorial}.

@@

The mono audio stream is located between 30~Hz and 15~kHz.
This signal is built by the sum of the left and right audio channels.
%The lower limit at 30~Hz is to prevent the transmission of a direct current (DC) part, which would require a large amount of power in transmission.
%This effect is one of the downsides of amplitude modulation (AM), where the carrier is located at the channels' center frequeny and requires a high transmission power. % TODO: check this fact
The upper limit at 15~kHz is chosen to maintain a sufficient spacing to the first subcarrier.
For audio streaming, as in FM broadcasting, this is not a limitation, since it is unlikely to have frequencies higher than 15~kHz in an audio stream.
Also, this already reaches the upper limit of the human ears' bandwidth, so higher spectral parts will not be heard anyway.

@@

The first subcarrier is allocated at an offset of 19~kHz from the center frequency.
This subcarrier is also called the pilot tone, since it is a continuous signal.
It is independent of any varying message signal content.
The pilot tone is used for stereo audio demodulation.
To regenerate a stereo audio signal, the left and right audio channels need to be recovered correctly.

@@

A signal that is constructed by the difference between the left and right audio channel is modulated on a 38~kHz subcarrier.
This subcarrier is an integer multiple of the pilot tone at 19~kHz, for practical reasons.
However, the 38~kHz carrier is suppressed and thus not visible in the received spectrum.
The modulation technique that is used for this spectral part is called dual-sideband suppressed-carrier (DSB-SC).
Even though the carrier is suppressed, it can still be recovered, because it is phase coherent with the 19~kHz pilot tone per definition.\\

The bandwidth for this difference-signal spans 15~kHz on either side of the subcarrier.
It is used to generate a stereo audio signal, in combination with the mono signal.
This process is explained into more detail in Chapter \ref{sec:demodulation-of-broadcast-fm}.

@@

Considering all these parts in the spectrum, there is still free bandwidth available to use up to the maximum channel bandwitdh of 100~kHz in one sideband.
Because of that, additional services were added to the pure audio transmission, to provide additional data services and information.\\

Services that were implemented are the Data Radio Channel (DARC), which is mostly used in Japan and the USA, the Subsidary Communication Authorization (SCA) and the Radio Data System (RDS) \cite{RohdeSchwarzFM}.
Out of these, RDS is the most significant service in Europe.
It is used to transmit additional information about the channel, such as the radio stations' names, the currently playing song's title or traffic information.

@@

The differentiator already converts the FM signal into an AM signal, but still leaves an FM content.
The transformation that happens here can be expressed in a formula, by simply differentiating Equation \eqref{equ:fm_func}.
This operation, with some additional simple algebraic transformations, results in
\begin{equation}
  \frac{d s(t)}{dt} = A_c 2 \pi \Big[f_c + k_f m(t) \Big] sin \Big(\omega_c t + 2 \pi k_f \int m(t) dt -pi \Big)
  \label{equ:fm_demod_discriminator}
\end{equation}

This formula shows AM and FM contents clearly.
The FM part is described by the sine function, with its argument.
The more important part is the AM part, which resembles the message signal.
It is described by the term within the square brackets.

@@

An important requirement for the differentiator is, that the input signal is of a constant amplitude \cite{SchnyderHaller2002}.
The transmitted signal is subject to random additive noise on the transmission channel.
The differentiator would deliver wrong results because of this noise, when substracting two consecutive samples, as described in Equation \eqref{equ:differentiator_two_samples}.
To achieve a constant amplitude, the received complex baseband signal $s(t)$ can be normalized to a value of one.
\begin{equation}
  s_{norm} = \frac{s}{|s|} = \frac{A(n)\ e^{j\phi_{\text{FM}}(n)}}{|{A(n)\ e^{j\phi_{\text{FM}}(n)}|}} = e^{j\phi_{\text{FM}}(n)}
\end{equation}
\begin{equation}
  |e^{j\phi_{FM}(n)}| = 1
\end{equation}

@@

The differentiated signal still consists of a FM part, as explained above, using Equation \eqref{equ:fm_demod_discriminator}.
In order to remove this high-frequeny part and extract the low-frequency envelope, the signal needs to be fed into an envelope detector.
A block diagram is shown in Figure \ref{fig:bd_freq_discriminator}.\\

@@

The first part of the channel frequency spectrum is the mono audio signal.
It is the summation signal of the left and right audio channel.
Thus, it contains enough information to replay an audio stream in mono quality.
A mono receiver can thus simply replay the so-called multiplex signal that is produced by the FM demodulator, without any further filtering or processing.
The 19~kHz pilot tone, and any higher frequency components, will not be audible by most people, because it is outside of the range of a human ear.
Besides that, any used speaker may also be unable to produce a frequency in this range.\\

Consequently, in order to demodulate this spectral part, a lowpass filter is required.
The filter needs to have a cutoff frequency of 15~kHz, to allow the entire mono audio spectrum to pass.
Sufficient attenutation must be available before 19~kHz, in order to suppress the pilot tone.

@@

The block diagram in Figure \ref{fig_bd_stereo_demod} illustrates the signal processing chain for stereo audio.
In the block diagram, $x(t)$ represents the multiplex signal from the FM demodulator.
The signals $x_L(t)$ and $x_R(t)$ describe the left and right audio signal, respectively.\\

@@

The central branch selects the pilot tone at 19~kHz with a bandpass filter.
This bandpass needs to have a frequency response that is sharp enough to have a sufficient attenuation at $\pm$4~kHz, since this is where the mono and difference signals' frequency spectra are allocated.
Afterwards, the pilot tone frequency is doubled to achieve a 38~kHz subcarrier.
This guarantees a phase-coherency between pilot and subcarrier, which is required to correctly shift the difference signal down to baseband.\\

The frequency doubling can be implemented with multiple methods.
The pilot tone can be modulated by simply multiplying it with itself.
Another option is to use a PLL and configure it to generate the double frequency at its output.
The PLL lock indicator can then be re-used as an indicator, whether a pilot tone exists.\\

In the upper branch, a bandpass filter selects the difference signal with ranges from 23 to 53~kHz.
It is then modulated to baseband by the previously generated 38~kHz subcarrier.
Subsequently, a lowpass limits signal bandwidth to 15~kHz, which removes the generated modulation artifacts.\\

@@

The rightmost part in the diagram performs the combination of the mono and the difference signal, according to Equation \eqref{equ_stereo_from_sum_diff}.
